---
title: "Getting Started with Benvos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Benvos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette serves to introduce the philosophy and structure of the `rbenvo` package, which implements the `benvo` class where `benvo` standards for Built Environment Object. This class makes working with built environment data easier by offering a convenient interface and methods for working with this type of [relational data](https://r4ds.had.co.nz/relational-data.html). We'll illustrate some of the very basics of how to use benvo's in this vignette as well as give some references as to how it might be used in other packages or settings.

We'll begin by loading the library and the two example data sets involving subjects living near Fast Food Restaurants (FFR). One of the two data sets will be for the subject level information, `FFR_subjects`, the other 
contains some measure of exposure (Distance/Time) that describe how long and/or how close subjects are located near the Built Environment Features (BEFs).

```{r setup,message=F,results=F}
library(rbenvo)
library(ggplot2)
theme_set(theme_bw())
data("FFR_distances")
data("FFR_subjects")
```


The subject dataframe should look familiar. This is a tidy dataset with an ID and two subject-level measures: BMI and sex.

```{r subjdf}
head(FFR_subjects,5)
```

The distance dataframe is a tidy "long" dataframe, containing multiple nearby BEF exposures per subject. Note that there *must be* at least two columns in every BEF dataset added to a benvo. One for the id and one for the Distance/Time data measured. Additionally, the latter columns must be named as "Distance" or "Time",as benvo's use these titles to perform their methods.
```{r befdf}
head(FFR_distances,5)
```

In order to create a benvo we'll pass the two dataframes to the `benvo()` function, wrapping the FFR_distances data frame in a list. 

```{r mkbenvo}
bdf <- benvo(subject_data = FFR_subjects,
             sub_bef_data = list(FFR=FFR_distances),by='id')
summary(bdf)
```

And that's it! That's all it takes to create a benvo, which is very particular kind of wrapper function for handling this specific kind of relational data in which there is a [one-to-many](https://en.wikipedia.org/wiki/One-to-many_(data_model)) relationship between the subject data frame and each BEF data frame.


## Benvo Methods

While the summary method above gives us a look at the overall `benvo` object, if we simply print the `benvo`, we'll get a view of whatever data frame is *active*.
```{r summary}
bdf
```

This notion of an *active* dataframe is inspired by (or borrowed from, depending on your perspective), the set-up of the [tidygraph](https://github.com/thomasp85/tidygraph/) package which also has a particular kind of relational data structure.

Essentially, the *active* dataframe is that dataframe which is currently available for viewing and editing via the popular [dplyr](https://dplyr.tidyverse.org/) verbs.


```{r verb_demo}
bdf %>% activate(FFR) %>% filter(id<=5)
```


There are also a number of helpful plotting functions specific to built environment data built in, as one last descriptive offering, we can plot a point-range of the distances for each subject using the 
`plot` function and passing the BEF label we want plotted. You'll see that there are several observations missing, since not every subject lives near a FFR.


```{r plot}
plot(bdf)
```


# Other packages

While the above descriptive statistics are nice, the real power of using `benvos` is that they offer 
functions like `joinvo` to easily join the subject and bef data frames, which can be useful in setting up the data for models in other packages. Below we'll use the `joinvo` function, passing the benvo and bef character label for which table we want joined.

```{r}
joinvo(bdf,"FFR")
```

Similarly, it is often the case - in my work at least - that some sort of model involving both the built environment data and the subject level data is needed. This means that information is incorporated at the subject level. This could manifest in, for example, the construction of 
a subject design matrix. Below I'll demonstrate how this works in a cross sectional as well as longitudinal framework, whilst highlighting how some of the previous functions take advantage
of the longitudinal structure.

```{r subjectdesign_example}
str(subject_design(bdf,BMI ~ sex))
```

### Longitudinal Example

For our longitudinal example we'll look at a dataset that imagines a hypothetical set of grocery or "Healthy Food Stores"(HFS)  near subjects. Subjects or grocery stores may move over
time so this will be reflected in our data. Below we'll repeat the similar descriptives from above, highlighting the differences in a longitudinal `benvo`.
```{r longitudinal_example}
data("longitudinal_HFS")
longitudinal_HFS
```

```{r longitudinal_summary}
summary(longitudinal_HFS)
```


```{r long_plot}
plot(longitudinal_HFS,
     plotfun = "pointrange",
     term ="HFS",
     component = "Distance",
     p = 0.9)
```

Analogous to `subject_design` there is also the `longitudinal_design` function which effectively allows for `lme4` style formulas to be used, incorporating subject or group level effects from the `subject_data` in the `benvo` and returning a `glmod` object along with the corresponding `X` and `y` design matrix and outcome vector, respectively.

```{r longitudinal_design_ex}
str(longitudinal_design(longitudinal_HFS,BMI ~ sex + (1|id)),max.level=1)
```




## Further Information

This completes this preliminary introduction to the `rbenvo` package. For current uses please see the [`rsstap`](https://apeterson91.github.io/rsstap/) or  [`rstapDP`](https://apeterson91.github.io/rstapDP/) packages. I've found that having these data structures greatly simplifies my code and allows for data to be organized more easily. If you're working with similar data, I hope you find the same!


